Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-11-14 Scripts/Manuscript.Rmd
Addressed comments for first round

---
title: Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  word_document: 
    reference_docx: ../References/template.docx
  pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
header-includes:
- \usepackage{amsmath}
---

@@

Over the last century, human activities have fundamentally changed the physico-chemical composition of the ocean, substantially warming the waters, decreasing the pH and reducing oxygen concentration [@IPCC:2019to]. Throughout the ocean, marine species are distributed according to preferences in given environmental parameters (e.g., salinity, temperature, oxygen) [@Hutchinson:1957nt; @Nelson:2016kk]. As a result of climate change, many marine species have changed their distribution towards higher latitude, deeper water or following local environmental gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions continue to change in the 21st century under climate change [@Cheung:2010dt]. Climate change is expected to impact fisheries production, compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Cheung:2016Sci; @Barange:2014kr; @Singht:2019]. The projected risks of impacts can be reduced through improving the effectiveness of current fisheries management [@Gaines:2018sg; @Cheung:2018dh], including managing fisheries on species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

@@

We propose that adapting the concepts of "Time of Emergence" (ToE) and “Thread Point” (TP) can help address some of the management challenges imposed by shifting species. Roughly, ToE is defined as the moment in time when an anthropogenic change signal (e.g., future trend) rises above the background noise of natural variability (e.g., historical variation) [@IPCC:2019tn]. This concept has been applied to investigate the emergence of multiple oceanic physical and biogeochemical variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind ToE is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background of natural climate variability [@Hawkins:2012hc]. The concept of thread point (TP) comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage shared stocks [@Sumaila:2013hv]. Specifically, Sumaila et al., *in press* applied this concept to understand the impacts of climate change to the economics of shared species between Canada and the United States.

@@

This analysis includes 633 marine exploited transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each species shared by a pair of neighbouring EEZs was considered a transboundary stock [@Teh:2015gd; @PalaciosAbrantes:2020tv]. We defined the boundaries of the world’s EEZs using the *Sea Around Us* spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided into Mexico Pacific and Mexico Atlantic; Figure \@ref(fig:FigureS1)) including island territories. We determined the intersections between polygons using the *R*-statistical software package *sf* [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by the geopolitical regions and natural biomes according to the United Nations and by natural biome [@Reygondeau:2019]. Habitat preference of each species was determined based on information available from FishBase (http://www.fishbase.org). For each species and EEZ, we used the *Sea Around Us* data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

We forced the DBEM with environmental data from a ten ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the ‘high’ climate change emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [Riahi:2011dk]. In GFDL-ESM2M, atmospheric surface temperature is projected to increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species’ distribution, we have to understand both the historical as well as the future climate-driven distribution variability, to infer differences between time frames. We do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M-ensembles were started from infinitesimally small differences in Earth system initial conditions resulting in a unique atmosphere and ocean state at each point in time. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

@@

We developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \ref{fig:FigureS2}). The centroid of a transboundary stock was determined by the average latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top $10^{th}$ percentile (e.g., *per* = 10%). We ran a sensitivity analysis (e.g. *per* = 20, 50, 90) to relax this parameter (Figure \ref{fig:FigureS2}). The centroid of a EEZ was estimated using the *st* package in *R* (version 3.2.1) (Figure \ref{fig:FigureS1}). For each ensemble member, neighboring EEZs and transboundary stock, we computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in *R*;

@@

\begin{equation}
\label{eq:two}

D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))

\end{equation}

@@

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a species and the centroid of the pair of neighbouring EEZs (A and B) sharing the species for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared species becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a state’s fisheries, higher TI indicates an increase in the challenges to the effective management and sustainability of the shared species’ fisheries. In contrast, a smaller TI suggests that the sharing of the species between neighbouring EEZs is becoming relatively more stable; a condition that favours effective management and sustainability of the shared species’ fisheries. The index was smoothed to a 10 year average to reduce interanual variabillity.

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than operating on a non-cooperative basis [@Munro:1980sd]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., *Threat Point* (TP) [@Sumaila:2013]. Here, we define TP as the minimum proportion of a shared stock within an EEZ that a country will accept to engage in cooperative management with the country’s sharing neighbour [@OCPalacios; @OCSumaila]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the shared stock.

First we estimated the stock share ratio (SSR) for each species and year (1951-2100) by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the species is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, we average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($t_h$) spans 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each species. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 (*te*), and mid $21^{st}$ century as the average of 2041 to 2060 (*tm*). We chose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected species distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_h$) following Palacios-Abrantes et al. (*in press*),

@@

Finally, the TP of each EEZ was calculated when $\Delta{SSR_{e,m}}$ is beyond one standard deviation of the historical distribution i.e., when $SSR_f \geq (SSR_f + \sigma)$ or $SSR_f \leq (SSR_f - \sigma)$.

@@

We tested the results for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@PackagegmtInterfa:2017vh] and *zoo* [@PackagezooSInfr:2019tm], *ggrepel* [@PackageggrepelAut:2020to], *zeallot* [@PackagezeallotMul:2018wy], *viridis* [@PackageviridisDef:2018us], *cowplot* [@CowplotStreamlined:2019wt], *moments* [@PackagemomensMome:2015uh] and *pgirmess* [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

### Time of Emergence of transbounudary species

We estimated the ToE between neighboring EEZs for 633 transboundary stocks, that is, the first year where the future stock’s distribution (2006-2100) overpasses the historical variation (1951-2005). Since 2006 till 2020, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least a set of neighboring EEZ (e.g., *Engraulis ringens* shared by Chile and Peru emerged in 2010). Stocks will continue to emerge steadily towards 2097 when all 633 species will have emerged in at least one neighboring EEZ (Figure \ref{fig:Figure1}). The number of EEZs with emerging stocks will increase steeply until 2025 when it seems to flatten approaching 2081 when all of the EEZs analyzed here will have at least one emerging stock (Figure \ref{fig:Figure1}).

```{r Figure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.cap =  "Cumulative increase in time of emergence by Exclusive Economic Zone and transboundary species in absolute numbers and percentage."}

@@

Most of the world EEZs are projected to experience an average ToE across transboundary species before 2040 (Figure \ref{fig:Figure2}). In a few EEZs species are projected to have a relative late average ToE (e.g. Honduras average ToE is 2071) while 54 of the 280 EEZs (19%) have already seen their species emerged (e.g. average ToE across all species $\leq$ 2020). A higher number of emerging species will happen between latitudes 0$^o$ and 50$^o$N corresponding to regions of relatively small EEZs with multiple borders (e.g. The Caribbean and the Mediterranean) (Figure \ref{fig:Figure2}). In most cases, the current participation of the emerging species in a country’s total fishing revenue is less than 50% (Figure \ref{fig:Figure2}- Polygons). However, in some cases emerging species represent 85% of current transboundary fishing revenue (e.g. Peru) or more (e.g., Marshall Islands, 92%). Only in very few cases like Brazil and Alaska’s polar EEZ, no stock will emerge between 2006 and 2100.

```{r Figure2, eval = T,  echo = F, fig.width  = 13, fig.height =11, fig.cap =  "Average time of emergence of transboundary specie. Land polygon shows the vulnerability index. Exclusive Economic Zone polygon represents the average time of emergence across all species. Density plot on the right shows the latitudinal number of emerging species across EEZ centroids (10^o running mean)."}

@@

Regionally, most of African, Asian and European EEZs are projected to have a later ToE, while the Americas and Oceania will see an early ToE (Figure \ref{fig:Figure3}). Twelve of the 47 countries responsible for 75% of the yearly revenue generated from transboundary species between 2005 and 2010 will see an average ToE before 2030 (Figure \ref{fig:Figure3}). This includes the United States who occupies the $4^{th}$ place in revenue generated from transboundary fisheries. These 47 countries are also among those with the largest number of emerging taxa (Figure \ref{fig:Figure3}). From the top ten countries with more revenue from transboundary species, nine of them share a border. Most of these countries (e.g. China) will also have a large number of emerging stocks, while others like Peru and Norway will have a small number of emerging taxa but representing high revenue fisheries.

```{r Figure3, eval = T,  echo = F, fig.width=4, fig.height=5, fig.cap =  "Average time of emergence of transboundary species per fishing entity. Fishing revenue in logarithmic scale. Showing names for the top 75th percentile. Countries marked with an asterisk have at least one neighboring border in the group."}

@@

We compared species ToE within EEZs by geographic region and by species habitat preference (Figure \ref{fig:Figure4}). The median ToE of transboundary species varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, *DF* = 93, *p* < 0.001). Post-hoc tests revealed regional patterns. EEZs in Easter Asia which have the latest average ToE across all habitat types (2041), and Latin America and the Caribbean and Polynesia who have the earliest average ToE across all habitats (2021 and 2018, respectively), significantly differ from (almost) all other regions. Other regions also significantly differ from each other but in a less frequent manner (see Table \ref{tab:TableS1}). When looking specifically at habitat association, we found significant differences in ToE by habitat preference (Kruskal-Wallis, $X^2$ = 286.48, *DF* = 93, *p* < 0.001). Specifically, the ToE of species with preferences for pelagic-oceanic and reef habitats was significantly different than other habitat preferences (See Table \ref{tab:TableS2}).

```{r Figure4, eval = T,  echo = F, fig.width=14, fig.height=14, fig.cap =  "Regional time of emergence per habitat preference. Regions according to the United Nations and habitat preference to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers outside of 1.5*IQR."}

@@

### Changes in specie Share Ratio (SSR)

@@

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings here are noteworthy in several ways. First, we show that the current distribution of most transboundary species have already emerged from the historical record and that virtually all EEZs have seen at least one emerging species since 2006 (Figure \ref{fig:Figure5}). Second, we identify regional patterns of complex changes in SSR and levels of vulnerability of transboundary fisheries management to climate change (Figure \ref{fig:Figure3}). Third, we identified that the historical share proportion of transboundary species will change by 2030 in most neighboring EEZs (Figure \ref{fig:Figure5}).

#### Emergence characteristics of transboundary species

>>>>

---
title: Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  # word_document: 
    # reference_docx: ../References/template.docx
  pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

@@

Over the last century, human activities have fundamentally changed the physico-chemical composition of the ocean, substantially warming the waters, decreasing the pH and reducing oxygen concentration [@IPCC:2019tn]. Throughout the ocean, marine species are distributed according to preferences in given environmental parameters (e.g., salinity, temperature, oxygen) [@Hutchinson:1957nt; @Nelson:2016kk]. As a result of climate change, many marine species have changed their distribution towards higher latitude, deeper water or following local environmental gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions continue to change in the 21st century under climate change [@Cheung:2010dt]. Climate change is expected to impact fisheries production, compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Barange:2014kr; @Singh:2017ds]. The projected risks of impacts can be reduced through improving the effectiveness of current fisheries management [@Gaines:2018sg; @Cheung:2018dh], including managing fisheries on species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

@@

We propose that adapting the concepts of "Time of Emergence" (ToE) and “Thread Point” (TP) can help address some of the management challenges imposed by shifting species. Roughly, ToE is defined as the moment in time when an anthropogenic change signal (e.g., future trend) rises above the background noise of natural variability (e.g., historical variation) [@IPCC:2019tn]. This concept has been applied to investigate the emergence of multiple oceanic physical and biogeochemical variables [@Rodgers:2015jqq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind ToE is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background of natural climate variability [@Hawkins:2012hc]. The concept of thread point (TP) comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage shared stocks [@Sumaila:2013hv]. Specifically, Sumaila et al., *in press* applied this concept to understand the impacts of climate change to the economics of shared species between Canada and the United States.

@@

This analysis includes 633 marine exploited transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each time a species was shared by a pair of neighbouring EEZs it considered a transboundary stock [@Teh:2015gd; @PalaciosAbrantes:2020tv] for a total of 9,132 transboundary stocks. We defined the boundaries of the world’s EEZs using the *Sea Around Us* spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided into Mexico Pacific and Mexico Atlantic; Figure \@ref(fig:FigureS1)) including island territories. We determined the intersections between polygons using the *R*-statistical software package *sf* [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by the geopolitical regions and natural biomes according to the United Nations and by natural biome [@Reygondeau:2019uh]. Habitat preference of each species was determined based on information available from FishBase (http://www.fishbase.org). For each species and EEZ, we used the *Sea Around Us* data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

We forced the DBEM with environmental data from a ten ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jqq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the ‘high’ climate change emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [Riahi:2011dk]. In GFDL-ESM2M, atmospheric surface temperature is projected to increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species’ distribution, we have to understand both the historical as well as the future climate-driven distribution variability, to infer differences between time frames. We do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jqq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M-ensembles were started from infinitesimally small differences in Earth system initial conditions resulting in a unique atmosphere and ocean state at each point in time. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

@@

We developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref{fig:FigureS2}). The centroid of a transboundary stock was determined by the average latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top $10^{th}$ percentile (e.g., *per* = 10%). We ran a sensitivity analysis (e.g. *per* = 20, 50, 90) to relax this parameter (Figure \@ref{fig:FigureS2}). The centroid of a EEZ was estimated using the *st* package in *R* (version 3.2.1) (Figure \@ref{fig:FigureS1}). For each ensemble member, neighboring EEZs and transboundary stock, we computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in *R*;

@@

\begin{equation}
\label{eq:two}
D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))
\end{equation}

@@

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a species and the centroid of the pair of neighbouring EEZs (A and B) sharing the species for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared species becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a state’s fisheries, higher TI indicates an increase in the challenges to the effective management and sustainability of the shared species’ fisheries. In contrast, a smaller TI suggests that the sharing of the species between neighbouring EEZs is becoming relatively more stable; a condition that favors effective management and sustainability of the shared species’ fisheries. The index was smoothed to a 10 year average to reduce interanual variability.

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., *Threat Point* (TP) [@Sumaila:2013hv]. Here, we define TP as the minimum proportion of a shared stock within an EEZ that a country will accept to engage in cooperative management with the country’s sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the shared stock.

First we estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. We did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, we average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($t_h$) spans 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each stock. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 (*te*), and mid $21^{st}$ century as the average of 2041 to 2060 (*tm*). We chose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_h$) following Palacios-Abrantes et al. (*in press*),

@@

Finally, a substantial change in SSR of each EEZ was determined when $\Delta{SSR_{e,m}}$ was beyond one standard deviation of the historical distribution i.e., when $SSR_f \geq (SSR_f + \sigma)$ or $SSR_f \leq (SSR_f - \sigma)$. In other words, when the future SSR is below or above a neighbouring EEZ TP.

@@

We tested the results for normality (e.g. skewness = 1.02, kurtosis = 2.83) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@PackagegmtInterfa:2017vh] and *zoo* [@PackagezooSInfr:2019tm], *ggrepel* [@PackageggrepelAut:2020to], *zeallot* [@PackagezeallotMul:2018wy], *viridis* [@PackageviridisDef:2018us], *cowplot* [@CowplotStreamlined:2019wt], *moments* [@PackagemomensMome:2015uh] and *pgirmess* [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

### Time of Emergence of transboundary species

The average ToE of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years with the first stocks emerging in 2006 and until 2100 (Figure \@ref(fig:Figure1). In total, 5,745 stocks will emerge between 2006 and 2100, that is 63% of the total transboundary stocks considered in this study (9,132). About 55% (n = 3,154) of those stocks are projected to have their shifts in distributions emerged between 2006 and 2020. Also, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century and are not expected to stop there (Figure \@ref(fig:Figure1)). The number of EEZs with emerging stocks is projected to increase steeply from 87% to 96% during the period of 2006 to 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.74% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:cumulative)).

```{r cumulative, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Cumulative increase in time of emergence by Exclusive Economic Zone and transboundary species in absolute numbers and percentage."}

@@

The median ToE of transboundary stocks varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, X2 = 287.23, DF = 93, p < 0.001). Around one fifth (19%) of the EEZs have an average time of emergence across species prior to 2020 (Figure \@ref{fig:Figure2}). Notable exceptions where the ToE is relative late include Honduras (2071), New Zealand (2055), Ireland (2067) and Pitcairn (2070). The EEZs in Eastern Asia, Latin America and the Caribbean as well as Polynesia differed significantly (*p* < 0.05) from (almost) all other regions (see Table S1 for full statistics). Tropical latitudes also hold the majority of countries with high vulnerability to shifts in transboundary stocks due to climate change (Figure \@ref{fig:Figure2}). Only in very few cases like Brazil and Alaska’s and Russia’s polar EEZs, no stock will emerge between 2006 and 2100. From the top 10 more vulnerable countries, seven are located in tropical waters (Tokelau 100, the	Marshall Islands 92, Peru 88, Kiribati 80, Tuvalu 80, Papua New Guinea 78, and the Bahamas 74) while only three are outside the tropics (Estonia 86, Finland 82 and Sweden 78). We set an arbitrary threshold of  1 s.d. (67% confidence) to determine the ToE (see Methods). We tested the sensitivity of our results using a larger threshold (i.e.,  2 s.d for a 95% confidence). While the selected 1 s.d. threshold is less conservative than a 2 s.d., the distribution of results do not dramatically change between thresholds, with a difference in ToE of about 4 years (Figure S\@ref{fig:S2}).

```{r Figure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Average time of emergence of transboundary species. A) Land polygon shows the vulnerability index of individual country’s fisheries management to shifts in transboundary species. The warmer the colour the greater the vulnerability. Exclusive Economic Zone polygon represents the average time of emergence across all species, with the cooler the colour the later the average time of emergence. The Density plot on the right shows the latitudinal number of emerging species across EEZ centroids (10 degrees latitude running mean). B) Time of emergence according to the United Nations sub regions. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent points outside of 1.5*IQR."}

@@

Our analysis suggests that most countries responsible for 75% (n = 47) of the yearly revenue generated from transboundary species between 2005 and 2010 will see their stocks emerge in average before 2040 (\@ref{fig:Figure3}). In some cases like Indonesia, Thailand and the United States, stocks will emerge ass soon as 2030. These 47 countries are also among those with the largest number of emerging stocks. Notably, nine of the top ten countries earning the most revenue from transboundary species, share a border.

```{r Figure3, eval = T,  echo = F, fig.width=4, fig.height=5,fig.pos= "H", fig.cap =  "Average time of emergence of transboundary species per fishing entity, aggregated by colour according to region. Fishing revenue on a logarithmic scale. Showing country names for the top 90th percentile. Countries marked with an asterisk have at least one neighboring border in the group."}

@@

We compared species ToE within EEZs by species habitat preference according to FishBase's classification (Figure \@ref{fig:Figure4}). We found significant differences in ToE by habitat preference (Kruskal-Wallis, $X^2$ = 286.48, *DF* = 93, *p* < 0.001). Specifically, the ToE of species with preferences for pelagic-oceanic and reef habitats was significantly different than other habitat preferences (See Table \ref{tab:TableS2}). In average, species associated with pelagic oceanic such as tunas (*Thunnus sp.*) and bathy-pelagic habitats like blue whiting (*Micromesistius poutassou*) as well as species associated to corals (e.g., Bigeye snapper	*Lutjanus	lutjanus*) will have an earliest time of emergence (Figure \@ref{fig:Figure4}). Only species associated to bathydemersal habitats such as Alaska plaice	*Pleuronectes	quadrituberculatus* and other habitats will have an average ToE later than 2025 (Figure \@ref{fig:Figure4}).

```{r Figure4, eval = T,  echo = F, fig.width=14, fig.height=14,fig.pos= "H", fig.cap =  "Regional time of emergence per habitat preference. Regions according to the United Nations and habitat preference to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers outside of 1.5*IQR."}

@@

### Changes in the threat point of transboundary stocks

@@

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings highlight the early emergence of transboundary distribution shifts from historical variations for most of the shared stocks in the world indicating climate change is already affecting international fisheries management (Figure \@ref{fig:Figure1}). Moreover, we identify regional patterns of high climate risk to important fish stocks and associated fisheries at the present-day that will require fast responses from transboundary fisheries management (Figure \@ref{fig:Figure2} and \@ref{fig:Figure3}). Finally, we show that in many instances around the world, the incoming years will see many transboundary stocks shifting their shared proportion below acceptable levels for sharing nations potentially becoming future sources of conflict (Figure \@ref{fig:Figure5})

#### High present-day climate risk of on transboundary fisheries management
